\pkg{plotROC}: A Better Tool for Plotting ROC Curves

### 2015-01-29 Paper/sachs2014plotROC.Rmd
editing, moved some figures

---
title: '\pkg{plotROC}: An \proglang{R} Package to Improve the State of ROC Curve Plots in the Medical Literature'
shorttitle: '\pkg{plotROC}: An \proglang{R} Package for ROC Curve Plots'
plaintitle: 'plotROC: An R Package to Improve the State of ROC Curve Plots in the Medical Literature'
author:
- name: Michael C. Sachs
  affil: Biometric Research Branch, Division of Cancer Treatment and Diagnosis, National Cancer Institute
address:
- Michael C. Sachs
- 9609 Medical Center Drive, MSC 9735
- Bethesda, MD 20892
- "telephone: 240-276-7888"
email: michael.sachs@nih.gov
abstract: 'Plots of the receiver operating characteristic (ROC) curve are ubiquitous in medical research. Designed to simultaneously display the operating characteristics at every possible value of a continuous diagnostic test, ROC curves are used in oncology to evaluate screening, diagnostic, prognostic and predictive biomarkers. We evaluate current trends in the design of ROC curve plots with a literature review. Our review suggests that ROC curve plots are often ineffective as statistical charts and that poor design obscures the relevant information the chart is intended to display. We describe our new \proglang{R} package that was created to address the shortcomings of existing tools. The package has functions to create informative ROC curve plots, with sensible defaults, for use in print or as an interactive web-based plot. A web application was developed to reach a broader audience of scientists who do not use R.'
keywords: ROC curves; graphics; interactive; plots
bibliography: plotroc
...

@@

The Receiver Operating Characteristic (ROC) curve is used to assess the accuracy of a continuous measurement for predicting a binary outcome. In medicine, ROC curves have a long history of use for evaluating diagnostic tests in radiology and general diagnostics. ROC curves have also been used for a long time in signal detection theory.

The accuracy of a diagnostic test can be evaluated by considering the two possible types of errors: false positives, and false negatives. For a continuous measurement that we denote as $M$, convention dictates that a test positive is defined as $M$ exceeding some fixed threshold $c$: $M > c$. In reference to the binary outcome that we denote as $D$, a good outcome of the test is when the test is positive among an individual who truly has a disease: $D = 1$. A bad outcome is when the test is positive among an individual who does not have the disease $D = 0$.

Formally, for a fixed cutoff $c$, the true positive fractions is the probability of a test positive among the diseased population:

@@

and the false positive fraction is the probability of a test positive among the healthy population:

@@

Since the cutoff $c$ is not usually fixed in advance, we can plot the TPF against the FPF for all possible values of $c$. This is exactly what the ROC curve is, $FPF(c)$ on the $x$ axis and $TPF(c)$ along the $y$ axis. A useless test that is not informative at all in regards to the disease status has $TPF(c) = FPF(c)$ for all $c$. The ROC plot of a useless test is thus the diagonal line. A perfect test that is completely informative about disease status has $TPF(c) = 1$ and $FPF(c) = 0$ for all $c$.

@@

A common way to summarize the value of a test for classifying disease status is to calculate the area under the ROC curve (AUC). The greater the AUC, the more informative the test. The AUC summarizes the complexities of the ROC curve into a single number and therefore is widely used to facilitate comparisons between tests and populations. It has been criticized for the same reason because it does not fully characterize the trade-offs between false- and true-positives.

@@

The main purpose of visually displaying the ROC curve is to show the trade-off between the FPF and TPF as the cutoff $c$ varies. This can be useful for aiding viewers in choosing an optimal cutoff for decision making, for comparing a small number of candidate tests, and for generally illustrating the performance of the test as a classifier. In practice, once the FPF and TPF are computed for each unique observed cutoff value, they can be plotted as a simple line chart or scatter plot using standard statistical plotting tools. This often leads to the unfortunate design choice of obscuring the critical and useful third dimension, the range of cutoff values $c$.

Another key design element is the use of a diagonal guideline for comparison. The diagonal guideline serves as a baseline for comparison, allowing observers to roughly estimate the area between the diagonal and the estimated ROC curve, which serves as a proxy for estimating the value of the test for classification above a coin-flip. Likewise, gridlines inside the plotting region and carefully selected axes allow for accurate assessment of the TPF and FPF at particular points along the ROC curve. Many medical studies use ROC curves to compare a multitude of candidate tests to each other and to the null diagonal. In those cases, curves need to be distinguished by using a legend combined with different colors or line types, or direct labels inside the plotting region.

In the medical literature, FPF and TPF are usually referred to in terms of the jargon terms Sensitivity and Specificity. Sensitivity is equivalent to the true positive fraction. Specificity is 1 - FPF, the true negative fraction. Sometimes, the FPF and TPF are incorrectly referred to as rates, using the abbreviations FPR and TPR. These are probabilities and their estimates are proportions, therefore we prefer the use of the term fraction as opposed to rate.

@@

The ROC curve plot is, at the most basic level, a line graph. Therefore, once the appropriate statistics are estimated, existing plotting functions can be used to create an ROC curve plot. The addition of axis labels, titles, legends, and so on, indicate a chart as such. In our literature review, we observed plots with the distinctive characteristics of the base plotting functions from Microsoft Office, \proglang{SAS}, \proglang{SPSS}, and the base \proglang{R} plotting functions.

There are several \proglang{R} packages related to ROC curve estimation that contain dedicated plotting functions. The \pkg{ROCR} package [@rocr] plots the FPF versus TPF, as usual, and then takes the interesting approach of encoding the cutoff values as a separate color scale along the ROC curve itself. A legend for the color scale is place along the vertical axis on the right of the plotting region. The \pkg{pROC} package [@pROC] is mainly focused on estimating confidence intervals and regions for restricted ranges of the ROC curve. The plotting methods therein use the base \proglang{R} plotting functions to create nice displays of the curves along with shaded confidence regions.

@@

Anyone giving a cursory look at any of the major medical journals is likely to find at least one ROC curve plot. We sought assess the usage of ROC curve plots and to evaluate the design choices made in the current oncology literature by conducting a small literature review. We searched Pubmed for  clinical trials or observational studies in humans reported in major oncology journals for the past 10 years for the terms "ROC Curve" OR "ROC Analysis" OR "Receiver operating characteristic curve". The search was conducted on October 8, 2014 and returned 54 papers. From those papers, 47 images were extracted and reviewed. The exact specifications for the Pubmed query are available in the supplementary materials.

Each image consisted of a single ROC curve plot or a panel of multiple plots. Each plot was inspected manually for the following design features: the number of curves displayed, the type of axis labels (sensitivity/ 1 - specificity or true/false positive fractions), presence or absence of grid lines, presence or absence of diagonal guide line, whether any cutpoint were indicated, the type of curve label (legend or direct label), and presence of other textual annotations such as the area under the curve. The numerical results of the survey are summarized in table \ref{table1}.

@@

rev <- read.csv("../../Literature Review/rocusage.csv", stringsAsFactors = FALSE)
rev <- subset(rev, flag != "comedy")
rev$axis.names <- gsub("fpr/tpr", "FPF/TPF", rev$axis.names, fixed = TRUE)

@@

The panels of figure \ref{figure1} illustrate the most common styles of ROC curve plots, and the associated design elements. We favor the use of gridline and reference lines to facilitate accurate readings off of the axes. Direct labels are preferred over legends because they omit the additional cognitive step of matching line types to labels. Our \pkg{plotROC} package additionally provides plotting of cutoff values, using hover events in interactive use, and direct labels for print use. Exact confidence regions for points on the ROC curve are optionally calculated and displayed. Additionally, we use axis scales are adjusted to be denser near the margins 0 and 1. In medical applications, it is often necessary to have a very low FPR (less than 10%, for instance), therefore the smaller scales are useful for accurately determining values near the margins. The next section details the usage of the \pkg{plotROC} \proglang{R} package and these features.

@@

# A
plot(TPF ~ FPF, data = R1, type = 's', xlab = "1 - Specificity", ylab = "Sensitivity", main = "A")
# B
plot(TPF ~ FPF, data = R1, type = 's', xlab = "1 - Specificity", ylab = "Sensitivity", main = "B")
lines(TPF ~ FPF, data = R2, type = 's', lty = 2)
legend("bottomright", legend = c("Test A: AUC = 0.81", "Test B: AUC = 0.64"), lty = 1:2)

@@

@@

```{r dataset}
library(plotROC)
D.ex <- rbinom(100, size = 1, prob = .5)
M.ex <- rnorm(100, mean = D.ex)
```

Next we use the `calculate_roc` function to compute the empirical ROC curve. The disease status need not be coded as 0/1, but if it is not, `plotROC` assumes (with a warning) that the lowest value in sort order signifies disease-free status. This returns a dataframe with three columns: the cutoff values, the TPF and the FPF.

@@

\begin{figure}[ht]
\centering
\includegraphics{screen-shot.pdf}
\caption{Screen shot of an interactive plot created with \pkg{plotROC} being displayed in the Rstudio viewer. Hovering the mouse cursor over the plot causes the cutoff label nearest to the cursor to be displayed. Clicking will display a confidence region, if available, and make the label stick until the next click. For live examples, see the package vignette, or go to http://sachsmc.github.io/plotROC. \label{interact}}
\end{figure}

@@

```{r print2, fig.width = 6, fig.height = 6, fig.cap = "Using \\pkg{ggplot2} themes and annotations with \\pkg{plotROC} objects. \\label{annotate}"}
library(ggplot2)
plot_journal_roc(myrocplot, rocdata) + 
  theme_grey() + 
  geom_abline(intercept = 0, slope = 1, color = "white") + 
  ggtitle("Themes and annotations") + 
  annotate("text", x = .75, y = .25, 
           label = "AUC = 0.80") + 
  ylab("Sensitivity")
```

@@

@@

\begin{figure}[ht]
\centering
\includegraphics{diagram.pdf}
\caption{Flowchart illustrating the approach that \pkg{plotROC} takes to generate either static plots for print or interactive plots for web-use. \label{flow}}
\end{figure}

>>>>

---
title: '\pkg{plotROC}: A Better Tool for Plotting ROC Curves'
shorttitle: '\pkg{plotROC}: A Better Tool for Plotting ROC Curves'
plaintitle: 'plotROC: A Better Tool for Plotting ROC Curves'
author:
- name: Michael C. Sachs
  affil: Biometric Research Branch, Division of Cancer Treatment and Diagnosis, National Cancer Institute
address:
- Michael C. Sachs
- 9609 Medical Center Drive, MSC 9735
- Bethesda, MD 20892
- "telephone: 240-276-7888"
email: michael.sachs@nih.gov
abstract: 'Plots of the receiver operating characteristic (ROC) curve are ubiquitous in medical research. Designed to simultaneously display the operating characteristics at every possible value of a continuous diagnostic test, ROC curves are used in oncology to evaluate screening, diagnostic, prognostic and predictive biomarkers. We reviewed a sample of ROC curve plots from the major oncology journals in order to assess current trends in usage and design elements. Our review suggests that ROC curve plots are often ineffective as statistical charts and that poor design obscures the relevant information the chart is intended to display. We describe our new \proglang{R} package that was created to address the shortcomings of existing tools. The package has functions to create informative ROC curve plots, with sensible defaults, for use in print or as an interactive web-based plot. A web application was developed to reach a broader audience of scientists who do not use R.'
keywords: ROC curves; graphics; interactive; plots
bibliography: plotroc
...

@@

The Receiver Operating Characteristic (ROC) curve is used to assess the accuracy of a continuous measurement for predicting a binary outcome. In medicine, ROC curves have a long history of use for evaluating diagnostic tests in radiology and general diagnostics. ROC curves have also been used for decades in signal detection theory.

The accuracy of a diagnostic test can be evaluated by considering two possible types of errors: false positives, and false negatives. For a continuous measurement that we denote as $M$, convention dictates that a test positive is defined as $M$ exceeding some fixed threshold $c$: $M > c$. In reference to the binary outcome that we denote as $D$, a good outcome of the test is when the test is positive for an individual who truly has a disease: $D = 1$. A bad outcome is when the test is positive for an individual who does not have the disease $D = 0$.

Formally, for a fixed cutoff $c$, the true positive fraction is the probability of a test positive in the diseased population:

@@

and the false positive fraction is the probability of a test positive in the healthy population:

@@

Since the cutoff $c$ is not fixed in advance, we can plot the TPF against the FPF for all possible values of $c$. This is exactly what the ROC curve is, a plot of $FPF(c)$ on the $x$ axis and $TPF(c)$ along the $y$ axis as $c$ varies. A useless test that is not informative at all in regards to the disease status has $TPF(c) = FPF(c)$ for all $c$. The ROC plot of a useless test is thus the diagonal line. A perfect test that is completely informative about disease status has $TPF(c) = 1$ and $FPF(c) = 0$ for all $c$.

@@

A common way to summarize the value of a test for classifying disease status is to calculate the area under the ROC curve (AUC). The greater the AUC, the more informative the test. The AUC summarizes the complexities of the ROC curve into a single number and therefore is widely used to facilitate comparisons between tests and across populations. It has been criticized for the same reason because it does not fully characterize the trade-offs between false- and true-positives.

@@

The main purpose of visually displaying the ROC curve is to show the trade-off between the FPF and TPF as the cutoff $c$ varies. This can be useful for aiding viewers in choosing an optimal cutoff for decision making, for comparing a small number of candidate tests, and for generally illustrating the performance of the test as a classifier. In practice, once the FPF and TPF are computed for each unique observed cutoff value, they can be plotted as a simple line chart or scatter plot using standard plotting tools. This often leads to the unfortunate design choice of obscuring the critical and useful third dimension, the range of cutoff values $c$.

Another key design element is the use of a diagonal guideline for comparison. They allow observers to roughly estimate the area between the diagonal and the estimated ROC curve, which serves as a proxy for estimating the value of the test for classification above a useless test. Likewise, gridlines inside the plotting region and carefully selected axes allow for accurate assessment of the TPF and FPF at particular points along the ROC curve. Many medical studies use ROC curves to compare a multitude of candidate tests to each other. In those cases, curves need to be distinguished by using different colors or line types combined with a legend, or direct labels inside the plotting region.

In the medical literature, FPF and TPF are usually referred to in terms of the jargon sensitivity and specificity. Sensitivity is equivalent to the true positive fraction. Specificity is 1 - FPF, the true negative fraction. Sometimes, the FPF and TPF are incorrectly referred to as rates, using the abbreviations FPR and TPR. These are probabilities and their estimates are proportions, therefore we prefer the use of the term fraction as opposed to rate.

@@

The ROC curve plot is, at the most basic level, a line graph. Therefore, once the appropriate statistics are estimated, existing plotting functions can be used to create an ROC curve plot. Viewers can identify ROC plots through context, by observing the shape of the line, and through the addition of axis labels, titles, legends, and so on. In our review of the oncology literature, we observed plots with the distinctive characteristics of the plotting functions from Microsoft Office, \proglang{SAS}, \proglang{SPSS}, and the base \proglang{R} plotting functions [@arr].

There are several \proglang{R} packages related to ROC curve estimation that contain dedicated plotting functions. The \pkg{ROCR} package [@rocr] plots the FPF versus TPF, as usual, and then takes the interesting approach of encoding the cutoff values as a separate color scale along the ROC curve itself. A legend for the color scale is place along the vertical axis on the right of the plotting region. The \pkg{pROC} package [@pROC] is mainly focused on estimating confidence intervals and regions for restricted ranges of the ROC curve. The plotting methods therein use the base \proglang{R} plotting functions to create nice displays of the curves along with shaded confidence regions. My \pkg{plotROC} package uses the \pkg{ggplot2} [@ggplot2] plotting library to create clear, informative ROC plots, with interactive features for use on the web, and sensible defaults for use in print.

@@

Anyone giving a cursory look at any of the major medical journals is likely to find at least one ROC curve plot. We sought to assess the usage of ROC curve plots and to evaluate the design choices made in the current oncology literature by conducting a small literature review. We searched Pubmed for  clinical trials or observational studies in humans reported in major oncology journals for the past 10 years for the terms "ROC Curve" OR "ROC Analysis" OR "Receiver operating characteristic curve". The search was conducted on October 8, 2014 and returned 54 papers. From those papers, 47 images were extracted and reviewed. The exact specifications for the Pubmed query are available in the manuscript source files.

Each image consisted of a single ROC curve plot or a panel of multiple plots. Each plot was inspected manually for the following design features: the number of curves displayed, the type of axis labels (sensitivity/ 1 - specificity or true/false positive fractions), presence or absence of grid lines, presence or absence of a diagonal guide line, whether any cutpoints were indicated, the type of curve label (legend or direct label), and presence of other textual annotations such as the AUC. The numerical results of the survey are summarized in table \ref{table1}.

@@

rev <- read.csv("litreview/rocusage.csv", stringsAsFactors = FALSE)
rev <- subset(rev, flag != "comedy")
rev$axis.names <- gsub("fpr/tpr", "FPF/TPF", rev$axis.names, fixed = TRUE)

@@

The panels of figure \ref{figure1} illustrate the most common styles of ROC curve plots, and the associated design elements. We favor the use of gridline and reference lines to facilitate accurate readings off of the axes. Direct labels are preferred over legends because they omit the additional cognitive step of matching line types to labels. Our \pkg{plotROC} package additionally provides plotting of cutoff values, which are displayed interactively with the web-based output option, and direct labels for print use. Exact confidence regions for points on the ROC curve are optionally calculated and displayed. Additionally, we use axis scales are adjusted to be denser near the margins 0 and 1. In medical applications, it is often necessary to have a very low FPR (less than 10%, for instance), therefore the smaller scales are useful for accurately determining values near the margins. The next section details the usage of the \pkg{plotROC} \proglang{R} package and these features.

@@

# A
plot(TPF ~ FPF, data = R1, type = 's', xlab = "1 - Specificity", ylab = "Sensitivity", main = "A")
# B
plot(TPF ~ FPF, data = R1, type = 's', xlab = "1 - Specificity", ylab = "Sensitivity", main = "B")
lines(TPF ~ FPF, data = R2, type = 's', lty = 2)
legend("bottomright", legend = c("Test A: AUC = 0.81", "Test B: AUC = 0.64"), lty = 1:2, cex = .75)

@@

## Installation and loading

\pkg{plotROC} can be installed from the Comprehensive \proglang{R} Archive Network, or installed from source.

```{r load, eval = FALSE}
install.packages("plotROC")
library(plotROC)
```

@@

```{r dataset, echo = -1}
library(plotROC)
D.ex <- rbinom(100, size = 1, prob = .5)
M.ex <- rnorm(100, mean = D.ex)
```

Next we use the `calculate_roc` function to compute the empirical ROC curve. The disease status need not be coded as 0/1, but if it is not, `calculate_roc` assumes (with a warning) that the lowest value in sort order signifies disease-free status. This returns a dataframe with three columns: the cutoff values, the TPF and the FPF.

@@

\begin{figure}[ht]
\centering
\includegraphics{figure/screen-shot.pdf}
\caption{Screen shot of an interactive plot created with \pkg{plotROC} being displayed in the Rstudio viewer. Hovering the mouse cursor over the plot causes the cutoff label nearest to the cursor to be displayed. Clicking will display a confidence region, if available, and make the label stick until the next click. For live examples, see the package vignette, or go to http://sachsmc.github.io/plotROC. \label{interact}}
\end{figure}

@@

```{r print2, warning = FALSE, message = FALSE, fig.width = 6, fig.height = 6, fig.cap = "Using \\pkg{ggplot2} themes and annotations with \\pkg{plotROC} objects. \\label{annotate}"}
library(ggplot2)
plot_journal_roc(myrocplot, rocdata) + 
  theme_grey() + 
  geom_abline(intercept = 0, slope = 1, color = "white") + 
  ggtitle("Themes and annotations") + 
  annotate("text", x = .75, y = .25, 
           label = "AUC = 0.80") + 
  scale_y_continuous("Sensitivity") + 
  scale_x_continuous("1 - Specificity")
```

@@

## export as pdf

@@

\begin{figure}[ht]
\centering
\includegraphics{figure/diagram.pdf}
\caption{Flowchart illustrating the approach that \pkg{plotROC} takes to generate either static plots for print or interactive plots for web-use. \label{flow}}
\end{figure}
